---
title: "PHI ARFP 2023: Introduction to GIS"
author: "Jessica Godwin"
date: "June 22, 2023"
output: 
 html_document:
   toc: true
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'U:/PHI-ARFP-2023/Lectures/IntrotoGIS')

```

# Setup
## Load Libraries

```{r load_lib}
library(tidycensus)
library(sf)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
```

# GIS in R: `sf` and `sp`
Note: The following text was one of the first published for spatial data analysis in `R`, and relied on the `sp` package. 

Bivand, R. S., Pebesma, E. J., & Gomez-Rubio, V. (2008). *Applied spatial data analysis with R*. Springer.

In August, many parts of `sp` will be deprecated, and folks are moving to `sf`. Bivand and Pebesma have published an updated eBook here: [https://r-spatial.org/book/](https://r-spatial.org/book/)

`sf` cheetsheet here: [https://github.com/rstudio/cheatsheets/blob/main/sf.pdf](https://github.com/rstudio/cheatsheets/blob/main/sf.pdf)

## Spatial Data & `R` Objects

Spatial data can come in many forms, which are generalized into the following categories:

  * Points: GPS locations
  * Lines: Streets, rivers, transit lines
  * Polygons: Nations, states, counties, cities, Census tracts, lakes
  * Pixels (Rasters): Satellite imagine, remote sensing, continuous space
  
`sf` handles all of these fairly well, but in `sp` each type of data was given a separate object type in `R` (like the difference between a `matrix` and a `data.frame`).


# `tidycensus` and Polygons

  *  `R` Package by [Kyle Walker](https://github.com/walkerke).
    *  Tidy for engaging with US Census and ACS data
    * [Basic Use Vignette](https://walker-data.com/tidycensus/articles/basic-usage.html)
    *  Get [Census API key](http://api.census.gov/data/key_signup.html)  
  *  Book: [Analyzing US Census Data: Methods, Maps, and Models in R](https://walker-data.com/census-r/)
 

```{r loadtidy}
vignette("tidycensus")

## Load API key ####
source("U:/PHI-ARFP-2023/Lectures/IntrotoGIS/CensusAPIKey.R")
census_api_key(myKey) 

### Option to cache tidycensus shapefiles ####
options(tigris_use_cache = TRUE)
```

## Getting variable names
```{r tidycensus_vars}
?get_acs
acs_var_df <- load_variables(2021, "acs5", cache = TRUE)
View(acs_var_df)

?get_decennial
dec_var_df <- load_variables(2020, "pl", cache = TRUE)
View(dec_var_df)
```

## Creating a polygon with data

  * You can pull Census or ACS tables without geometry by specifying `geometry = FALSE`. 
  *  The `geography` argument has different options for specifications based on what year and data type you are requesting. 

```{r pull_census}
kingco_hu <- get_decennial(geography = "tract",
                           variables = c("H1_001N", "H1_002N",
                                         "H1_003N"),
                           state = "WA",
                           county = "King",
                           year = 2020,
                           sumfile = "pl",
                           geometry = TRUE)

View(kingco_hu)
summary(kingco_hu)
unique(kingco_hu$variable)
```

## Plot with `ggplot`

```{r tidycensus_plots}
kingco_hu %>% 
  ## We have 3 vars in long form data,
  ## i.e. each polygon has three rows in the data,
  ## let's filter down to total housing units
  filter(variable == "H1_001N") %>% 
  ggplot() +
  geom_sf(aes(fill = value)) 

kingco_hu %>% 
  filter(variable == "H1_001N") %>% 
  ggplot() +
  ## size makes polygon lines smaller
  geom_sf(aes(fill = value), size = .25) +
  ## Nice palette, and reverse direction
  ## so light is small, dark is large
  scale_fill_viridis_c(direction = -1,
                       name = "Units") + 
  ggtitle("King County Census Tracts (2020)") +
  ## git rid of long lat axes
  theme_void()
```

## Multiple variables with the same map

```{r facet_pivot}
kingco_hu %>% 
  ggplot() +
  geom_sf(aes(fill = value), size = .25) +
  scale_fill_viridis_c(direction = -1,
                       name = "Units") + 
  facet_wrap(~variable) +
  ggtitle("King County Census Tracts (2020)") +
  theme_void()

kingco_hu_wide <- kingco_hu %>% 
  pivot_wider(id_cols = c("GEOID", "NAME", "geometry"),
               names_from = "variable",
               values_from = "value")

range(kingco_hu$value)
hu_tot <- kingco_hu_wide %>% 
  ggplot() +
  geom_sf(aes(fill = H1_001N), size = 0.25) +
  scale_fill_viridis_c(direction = -1,
                       name = "Units",
                       breaks = c(0, 100, 500,
                                  1000, 1500,
                                  2000, 2500,
                                  3000, 3500,
                                  3750)) + 
  ggtitle("Total") +
  theme_void()

hu_occ <- kingco_hu_wide %>% 
  ggplot() +
  geom_sf(aes(fill = H1_002N), size = 0.25) +
  scale_fill_viridis_c(direction = -1,
                       name = "Units",
                       breaks = c(0, 100, 500,
                                  1000, 1500,
                                  2000, 2500,
                                  3000, 3500,
                                  3750)) + 
  ggtitle("Occupied") +
  theme_void()

hu_vac <- kingco_hu_wide %>% 
  ggplot() +
  geom_sf(aes(fill = H1_003N), size = 0.25) +
  scale_fill_viridis_c(direction = -1,
                       name = "Units",
                       breaks = c(0, 100, 500,
                                  1000, 1500,
                                  2000, 2500,
                                  3000, 3500,
                                  3750)) + 
  ggtitle("Vacant") +
  theme_void()

grid.arrange(hu_tot, hu_occ, hu_vac, nrow = 2)
```

# Coordinate reference systems and combining spatial data
 
  * ESRI (a company one of whose products is ArcGIS) shapefiles consist of at least three files, and this is a common form. The first file `.shp` contains the geography of each shape. The second file `shx` is an index file which contains record offsets. The third file `.dbf` contains feature attributes with one record per feature.
  * To combine multiple spatial objects, they must be on the same *coordinate reference system*.
  
Download King County school locations data from [https://gis-kingcounty.opendata.arcgis.com/datasets/school-sites-in-king-county-schsite-point?geometry=-123.799%2C47.157%2C-120.017%2C47.807](https://gis-kingcounty.opendata.arcgis.com/datasets/school-sites-in-king-county-schsite-point?geometry=-123.799%2C47.157%2C-120.017%2C47.807)
```{r download_points}
kc_schools <- read_sf("U:/PHI-ARFP-2023/Lectures/IntrotoGIS/School_Sites_in_King_County___schsite_point")

head(kc_schools)

st_crs(kc_schools)
st_crs(kingco_hu)

kc_schools <- kc_schools %>% 
  st_transform(crs = st_crs(kingco_hu))

st_crs(kc_schools)
```

## Plots with points and polygons

```{r 2sf}
kingco_hu_wide %>% 
  ggplot() +
  geom_sf(aes(fill = H1_001N), size = 0.25) +
  scale_fill_viridis_c(direction = -1,
                       name = "Units",
                       breaks = c(0, 100, 500,
                                  1000, 1500,
                                  2000, 2500,
                                  3000, 3500,
                                  3750)) + 
  geom_point(data = kc_schools, aes(x = LONG_CEN, y = LAT_CEN,
                                    color = "white")) +
  ggtitle("Total") +
  theme_void()
```


## Plots with two polygons

```{r 2sf2}
seattle <- read_sf("U:/PHI-ARFP-2023/Lectures/IntrotoGIS/City_of_Seattle_Shoreline/")

st_crs(seattle)

seattle <- st_transform(seattle, crs = st_crs(kingco_hu))

seattle_idx <- st_intersects(kingco_hu_wide, seattle)
head(seattle_idx)

king_idx_todrop <- lapply(seattle_idx, function(data_row){
  length(data_row) == 0
}) %>% unlist() %>% which()


kingco_hu_wide %>%
  slice(-king_idx_todrop) %>% 
  ggplot() +
  geom_sf(aes(fill = H1_001N), size = 0.25) +
  scale_fill_viridis_c(direction = -1,
                       name = "Units",
                       breaks = c(0, 100, 500,
                                  1000, 1500,
                                  2000, 2500,
                                  3000, 3500,
                                  3750)) +
  ggtitle("Total") +
  theme_void()
```
